Title

### 2012-10-31 inc/este/este.tex
add apl bibliography up to entry 13

%TODO consider moving this paragraph to chapter intro 
Sub-nanosecond (ultrafast) pulsed laser-driven electron sources are now being employed in a wide variety of scientific studies: in particular, ultrafast electron diffraction (UED) [1], ultrafast transmission electron microscopy (UEM) [2], and pulsed X-ray free electron lasers [3].
The spatial quality of these sources is often characterized in terms of their normalized transverse rms emittance [4,5]; $\varepsilon_{\smallT} = \Delta p_x \Delta x / (m_0 c)$, where $\Delta p_x = \sqrt{ \langle p_x^2 \rangle }$ is the rms transverse momentum of the source and $\Delta x = \sqrt{ \langle x^2 \rangle }$  is its size (cylindrical symmetry assumed), and $m_0$ is the electron rest mass and $c$ is the speed of light in vacuum.
As a reduction in the source size $\Delta x$ is limited either by the attainable diffraction-limited focusing or by the short-pulse Child's Law (screening of the electron gun's acceleration field) [6], significant improvements in the performance of scientific instruments employing pulsed electron sources through a reduction in their spatial emittance $\varepsilon_{\smallT}$ (and consequent increase in brightness [7]) may only be possible by decreasing the rms transverse momentum $\Delta p_x$ of the electron source --- an intrinsic property of the emission source [4,5].

In this section, we show that, at least for thermionic emission, the effective mass $m^*$ of the electronic state from which the electron was emitted affects the rms transverse momentum; that is, the expression for $\Delta p_x$ developed in Refs. 4 and 5 for thermionic electron emission should be rewritten to read $\sqrt{m^* k_B T_e}$, where $T_e$ is the temperature of the emitting electron distribution and $k_B$ is Boltzmann's constant.
Direct observation of the transverse momentum distribution for excited-state thermionic emission (ESTE) from GaSb and InSb, two similar III-V zincblende semiconductors, demonstrates the $m^*$ dependence.  A further experiment employing two-photon assisted thermionic emission (2PTE) from gold, for which $m^* = m_0$ [8], confirms the temperature dependence of $\Delta p_x$.  Simulations of the experiment using our extended analytical Gaussian (AG) model of electron pulse propagation (see Section \ref{sec:external_forces}) are in close agreement with the data.

Figure 1 depicts the experimental technique used to determine directly the transverse momentum distributions (and hence $\Delta p_x$) of the laser-driven electron sources.
The primary laser radiation source for the studies is a home-built, diode-pumped and thermal-lens-shaped, femtosecond Yb:KGW oscillator [10] delivering 250fs duration pulses at 1047nm and a 63MHz repetition rate (see Section \ref{sec:laser}).
This 2W average power laser is frequency doubled in a 3mm lithium triborate (LBO) crystal with 50\% efficiency to yield $\sim$200fs green pulses at 523nm (photon energy, $\hbar \omega$ = 2.37eV).
Further doubling with a cylindrical focusing geometry in a 7mm $\beta$-barium borate (BBO) crystal yields 261nm ($\hbar \omega$ = 4.75eV) UV pulses with a duration of $\sim$4ps (due to the 600fs/mm group velocity mismatch between the green and generated UV).
Both the green and UV laser pulses are focused with 30cm focal lenses onto the photocathode in a 20kV electron gun at an angle of incidence of $60(\pm5)^{\circ}$, which improves the coupling of the $p$-polarized laser radiation into the photocathode surface [7].
The electron gun design is based on the that of Togawa et al. [7,11] and has an acceleration gap of 25($\pm$2)mm.
After acceleration in the DC photo-gun, the electrons pass through a set of electrostatic deflection plates to ensure that the electron beam passes, as close as possible, axially through the center of two, 6.35mm-bore, thin and round magnetic lenses with counter-propagating currents to minimize image rotation effects.
A YAG scintillator placed 27.5cm after the second magnetic lens detects the spatial electron beam profile, and its visible fluorescence is 1:1 imaged onto a CCD detector with 5.4$\mu$m pixels.

@@

The quadratic dependence of the electron yield on pulse energy clearly indicates a two-photon emission process.
This is expected since an excitation energy of two green photons ($\hbar \omega$ = 2.37eV) will be required to overcome the reported $\phi$ = 4.69eV effective work function of a thick gold film contaminated with adsorbed water [12] or ensure that at least the tail of the electron Fermi distribution has sufficient energy to overcome the 5.1($\pm$0.1)eV work function of a clean Au surface [13].
Moreover, when $2\hbar \omega \approx \phi$, the dependence of the yield on the square of the electron's excess energy above the work function [12] ensures that predominantly only the high energy (Boltzmann) tail of the Fermi distribution contributes to the emission.
Consequently, the observed increase in $\Delta p_x$ with incident pulse energy must be due to a heating effect; specifically, laser heating of the free electron Fermi gas as two-photon excitation is instantaneous and the $\sim$200fs green laser pulse duration excludes any coupling to the lattice which occurs on the time scale of a few picoseconds [14].
The solid line in Figure 2 is the predicted variation derived using a simple zero free parameter model of this effect.
Using the known optical properties of gold [8], which give a reflectance $R_p$ = 0.53 and an absorption depth of 20nm, and the temperature-dependent heat capacity of a free electron Fermi gas, the model evaluates the average temperature of the laser-heated electron gas that may be two-photon excited assuming that only electrons within a few nanometers of the surface can be emitted, as the mean free path for electrons in Au is $\sim$4nm [15].
This average electron temperature $T_e$ is then used in the AG model simulation of the experiment (Figure 1) to determine the expected Fourier plane spot size on the YAG scintillator; that is, using an initial $\Delta p_x = \sqrt{m_0 k_B T_e}$.  The dashed lines indicate the expected trends for a $\pm$10\% error in the laser spot size $\Delta x_0$ incident on the gold cathode surface.
The close agreement between the data and the simulation strongly supports our interpretation of the emission mechanism and, as expected, that the effective mass of a free electron in Au is equal to the rest mass $m_0$ [8].

@@

Figure 3 also shows that the measured Fourier plane electron beam spot size for GaSb is independent of the UV laser pulse energy, as would be expected for the proposed emission process.
More significantly, using $\Delta p_x = \sqrt{m_0 k_B T_e}$ (i.e., assuming the emitted electrons have a mass $m_0$ in the semiconductor), simulation of the experiment with the extended AG model [9] indicates that the observed GaSb Fourier plane spot size of 0.45($\pm$0.03)mm (HWe$^{-1}$M) would be due to an electron temperature $T_e \approx$ 360K which is associated with negligible thermionic emission ($\exp[-\phi/(k_B T_e)] \sim 10^{-15}$).
On the other hand, employing a mass of 0.3$m_0$ for the electrons in GaSb allows an average electron temperature (over the $\sim$4ps laser pulse duration) of about 1,200K to be extracted using the AG model simulation --- a value much more consistent with the proposed ESTE mechanism and the expected cooling rate by LO phonon emission.
In fact, at the maximum $\sim$0.2nJ incident UV laser pulse energy up to 108 electrons/pulse could be excited into the upper $\Gamma_8$ conduction band, of which one in $\exp[\phi/(k_B T_e)] \approx$ 20,000 are above the effective 0.99eV work function, thus agreeing with the observed yield of $\sim$10$^3$ electrons/pulse (Figure 3).
We also note that the normalized transverse rms emittance $\varepsilon_{\smallT}$ of this UV laser-driven ultrafast GaSb electron source is more than a factor of two less than that expected from a standard Cu photocathode ($\phi$ = 4.31eV and assuming $\Delta p_x = \sqrt{m_0 ( \hbar \omega - \phi ) / 3 }$, where $\hbar \omega$ is the incident photon energy [4,5]) irradiated at the same 261nm wavelength.

>>>>

%TODO consider moving this paragraph to chapter intro 
Sub-nanosecond (ultrafast) pulsed laser-driven electron sources are now being employed in a wide variety of scientific studies: in particular, ultrafast electron diffraction (UED)\cite{sciaini_femtosecond_2011}, ultrafast transmission electron microscopy (UEM)\cite{king_ultrafast_2005}, and pulsed X-ray free electron lasers\cite{dowell_cathode_2010}.
The spatial quality of these sources is often characterized in terms of their normalized transverse rms emittance\cite{dowell_quantum_2009,jensen_emittance_2010}; $\varepsilon_{\smallT} = \Delta p_x \Delta x / (m_0 c)$, where $\Delta p_x = \sqrt{ \langle p_x^2 \rangle }$ is the rms transverse momentum of the source and $\Delta x = \sqrt{ \langle x^2 \rangle }$  is its size (cylindrical symmetry assumed), and $m_0$ is the electron rest mass and $c$ is the speed of light in vacuum.
As a reduction in the source size $\Delta x$ is limited either by the attainable diffraction-limited focusing or by the short-pulse Child's Law (screening of the electron gun's acceleration field)\cite{valfells_effects_2002}, significant improvements in the performance of scientific instruments employing pulsed electron sources through a reduction in their spatial emittance $\varepsilon_{\smallT}$ (and consequent increase in brightness\cite{berger_dc_2009}) may only be possible by decreasing the rms transverse momentum $\Delta p_x$ of the electron source --- an intrinsic property of the emission source\cite{dowell_quantum_2009,jensen_emittance_2010}.

In this section, we show that, at least for thermionic emission, the effective mass $m^*$ of the electronic state from which the electron was emitted affects the rms transverse momentum; that is, the expression for $\Delta p_x$ developed in Refs. 4 and 5 for thermionic electron emission should be rewritten to read $\sqrt{m^* k_B T_e}$, where $T_e$ is the temperature of the emitting electron distribution and $k_B$ is Boltzmann's constant.
Direct observation of the transverse momentum distribution for excited-state thermionic emission (ESTE) from GaSb and InSb, two similar III-V zincblende semiconductors, demonstrates the $m^*$ dependence.  A further experiment employing two-photon assisted thermionic emission (2PTE) from gold, for which $m^* = m_0$\cite{johnson_optical_1972}, confirms the temperature dependence of $\Delta p_x$.  Simulations of the experiment using our extended analytical Gaussian (AG) model of electron pulse propagation (see Section \ref{sec:external_forces}) are in close agreement with the data.

Figure 1 depicts the experimental technique used to determine directly the transverse momentum distributions (and hence $\Delta p_x$) of the laser-driven electron sources.
The primary laser radiation source for the studies is a home-built, diode-pumped and thermal-lens-shaped, femtosecond Yb:KGW oscillator\cite{berger_high-power_2008} delivering 250fs duration pulses at 1047nm and a 63MHz repetition rate (see Section \ref{sec:laser}).
This 2W average power laser is frequency doubled in a 3mm lithium triborate (LBO) crystal with 50\% efficiency to yield $\sim$200fs green pulses at 523nm (photon energy, $\hbar \omega$ = 2.37eV).
Further doubling with a cylindrical focusing geometry in a 7mm $\beta$-barium borate (BBO) crystal yields 261nm ($\hbar \omega$ = 4.75eV) UV pulses with a duration of $\sim$4ps (due to the 600fs/mm group velocity mismatch between the green and generated UV).
Both the green and UV laser pulses are focused with 30cm focal lenses onto the photocathode in a 20kV electron gun at an angle of incidence of $60(\pm5)^{\circ}$, which improves the coupling of the $p$-polarized laser radiation into the photocathode surface\cite{berger_dc_2009}.
The electron gun design is based on the that of Togawa et al.\cite{berger_dc_2009,togawa_ceb6_2007} and has an acceleration gap of 25($\pm$2)mm.
After acceleration in the DC photo-gun, the electrons pass through a set of electrostatic deflection plates to ensure that the electron beam passes, as close as possible, axially through the center of two, 6.35mm-bore, thin and round magnetic lenses with counter-propagating currents to minimize image rotation effects.
A YAG scintillator placed 27.5cm after the second magnetic lens detects the spatial electron beam profile, and its visible fluorescence is 1:1 imaged onto a CCD detector with 5.4$\mu$m pixels.

@@

The quadratic dependence of the electron yield on pulse energy clearly indicates a two-photon emission process.
This is expected since an excitation energy of two green photons ($\hbar \omega$ = 2.37eV) will be required to overcome the reported $\phi$ = 4.69eV effective work function of a thick gold film contaminated with adsorbed water\cite{monjushiro_ultraviolet_1991} or ensure that at least the tail of the electron Fermi distribution has sufficient energy to overcome the 5.1($\pm$0.1)eV work function of a clean Au surface\cite{eastman_photoelectric_1970}.
Moreover, when $2\hbar \omega \approx \phi$, the dependence of the yield on the square of the electron's excess energy above the work function\cite{monjushiro_ultraviolet_1991} ensures that predominantly only the high energy (Boltzmann) tail of the Fermi distribution contributes to the emission.
Consequently, the observed increase in $\Delta p_x$ with incident pulse energy must be due to a heating effect; specifically, laser heating of the free electron Fermi gas as two-photon excitation is instantaneous and the $\sim$200fs green laser pulse duration excludes any coupling to the lattice which occurs on the time scale of a few picoseconds [14].
The solid line in Figure 2 is the predicted variation derived using a simple zero free parameter model of this effect.
Using the known optical properties of gold [8], which give a reflectance $R_p$ = 0.53 and an absorption depth of 20nm, and the temperature-dependent heat capacity of a free electron Fermi gas, the model evaluates the average temperature of the laser-heated electron gas that may be two-photon excited assuming that only electrons within a few nanometers of the surface can be emitted, as the mean free path for electrons in Au is $\sim$4nm [15].
This average electron temperature $T_e$ is then used in the AG model simulation of the experiment (Figure 1) to determine the expected Fourier plane spot size on the YAG scintillator; that is, using an initial $\Delta p_x = \sqrt{m_0 k_B T_e}$.  The dashed lines indicate the expected trends for a $\pm$10\% error in the laser spot size $\Delta x_0$ incident on the gold cathode surface.
The close agreement between the data and the simulation strongly supports our interpretation of the emission mechanism and, as expected, that the effective mass of a free electron in Au is equal to the rest mass $m_0$ [8].

@@

Figure 3 also shows that the measured Fourier plane electron beam spot size for GaSb is independent of the UV laser pulse energy, as would be expected for the proposed emission process.
More significantly, using $\Delta p_x = \sqrt{m_0 k_B T_e}$ (i.e., assuming the emitted electrons have a mass $m_0$ in the semiconductor), simulation of the experiment with the extended AG model\cite{berger_semi-analytic_2010} indicates that the observed GaSb Fourier plane spot size of 0.45($\pm$0.03)mm (HWe$^{-1}$M) would be due to an electron temperature $T_e \approx$ 360K which is associated with negligible thermionic emission ($\exp[-\phi/(k_B T_e)] \sim 10^{-15}$).
On the other hand, employing a mass of 0.3$m_0$ for the electrons in GaSb allows an average electron temperature (over the $\sim$4ps laser pulse duration) of about 1,200K to be extracted using the AG model simulation --- a value much more consistent with the proposed ESTE mechanism and the expected cooling rate by LO phonon emission.
In fact, at the maximum $\sim$0.2nJ incident UV laser pulse energy up to 108 electrons/pulse could be excited into the upper $\Gamma_8$ conduction band, of which one in $\exp[\phi/(k_B T_e)] \approx$ 20,000 are above the effective 0.99eV work function, thus agreeing with the observed yield of $\sim$10$^3$ electrons/pulse (Figure 3).
We also note that the normalized transverse rms emittance $\varepsilon_{\smallT}$ of this UV laser-driven ultrafast GaSb electron source is more than a factor of two less than that expected from a standard Cu photocathode ($\phi$ = 4.31eV and assuming $\Delta p_x = \sqrt{m_0 ( \hbar \omega - \phi ) / 3 }$, where $\hbar \omega$ is the incident photon energy [4,5]) irradiated at the same 261nm wavelength.
